---
title: "EDA_Thesis"
author: "Yufang"
date: "2025-02-05"
output:
  pdf_document: default
  word_document: default
  html_document:
    df_print: paged
---
# 1. Exploratory Data Anlysis

## read in the data & clean the data with the same approach in Thei et al., 2024
```{r, echo=FALSE, warning=FALSE, results='hide'}
rm(list = ls())
library(dplyr)
library(survival)library(mstate)
library(ggplot2)
library(corrplot)
setwd("C:/Users/Alisa_Wang/Desktop/MasterThesis/Code" )
df <- readRDS("NOTR_DGF.rds")
sort(colnames(df))
```

```{r, echo=FALSE}
## INCLUSION & EXCLUSION CRITERIA
df_using <- subset(df,Typeofdonor=='Deceased' 
                      & Recipientage >17
                      & Donorage >17
                      & Number_of_transplantations==1
                      & (Combinedtransplants=='LKi'
                          | Combinedtransplants=='RKi') 
                      # & PNF== 0
                      # & is.na(Transatl) WHAT DOES THIS MEAN??
                      # & PT_Earlydeath ==0
                      # & !is.na(dDGF)
                      & Year_Of_Tx > 2013
                      & Year_Of_Tx < 2023)

```

# select the necessary variables
```{r, echo=FALSE, results='hide'}
variableName_df = read.csv(file = "../Data/codeboek2_.csv")
variableName = variableName_df$Name_NOTR 
variableName = if_else(variableName == "Initialweight", "Recipient_weight", variableName) # use the renamed name
variableName = if_else(variableName == "Initialheight", "Recipient_height", variableName) # use the renamed name
variableName = if_else(variableName == "Warmischaemicperiod1", "Warm_ischaemic_period_1", variableName) 
variableName = if_else(variableName == "InitialColdischaemicperiod", "Cold_ischaemic_period", variableName) 
# variableName = if_else(variableName == "Lowestcreatinine碌moll", "Lowestcreatinineµmoll", variableName) 

variableName_using = variableName[nzchar(variableName)]
variableName_using = variableName_using[!variableName_using %in% c("InitialPrimaryDisease", "InitialPrimaryDiseaserenine", "DonorHLAstring")]

variableName_using = c(variableName_using, c("time", "status", "status2", "HLA_mismatch")) # including survival related variables
df_using = df_using[variableName_using]

head(df_using)
str(df_using)
```


## replicate the visulized results
```{r, echo=TRUE}
sfall = survfit(Surv(time, status2 > 0) ~ 1, data = df_using)
plot(sfall, lwd = 2)

# cumulative incidences
ci = Cuminc("time", "status2", data = df_using)
head(ci)

plot(ci$time, ci$CI.1, type = "s", lwd = 2, col = "blue", ylim = c(0, 0.5),
     xlab = "Days since transplant", ylab = "Cumlative incidence")
lines(ci$time, ci$CI.2, type = "s", lwd = 2, col = "red")
legend("topleft", levels(df_using$status)[-1], lwd = 2, col = c("blue", "red"), bty = "n")
```

## summarize of data 1: to get the unqiue values for non-numeric & non-date variables 
```{r, echo=TRUE}
results <- data.frame(
  Variable = character(),
  LengthUniqueValues = integer(),
  MissingObs = integer(),
  UniqueValues = character()
)

# Loop through each column in the data frame
for (var in colnames(df_using)) {
  if (!is.numeric(df_using[[var]]) & !inherits(df_using[[var]], "Date")) {
    # Convert unique values to a single string
    unique_values_str <- paste(unique(df_using[[var]]), collapse = ", ")
    
    # Add the variable name, unique values, and number of unique values to the results data frame
    results <- rbind(results, data.frame(
      Variable = var,
      LengthUniqueValues = length(unique(df_using[[var]])),
      MissingObs = sum(is.na(df_using[[var]])),
      UniqueValues = unique_values_str
    ))
  }
}

# Print the results data frame
print(results)

# joint non-complete cases
sum(!complete.cases(df_using))
sum(complete.cases(df_using))
```

## deal with missing data issues
```{r, echo=TRUE}
# set the missing InitialOnmachineindicator being an additional category
df_using$InitialOnmachineindicator = if_else(is.na(df_using$InitialOnmachineindicator), "Unknown", df_using$InitialOnmachineindicator)

# impute the missing Smoking data
library(mice)
methods = make.method(df_using)
methods["Smoking"] = "pmm" # Predictive mean matching
methods[setdiff(names(df_using), "Smoking")]
# Perform the imputation with more iterations
imputed_df_using <- mice(df_using, method = methods, m = 5, maxit = 100, seed = 500)

# Check the imputed data
summary(imputed_df_using)

# Complete the dataset with the imputed values
completed_df_ <- complete(imputed_df_using)

# group the InitialPrimaryDiseaseET


```


### re-label/factorization variables which have 2 or three unique values
```{r, echo=FALSE}
df_using = df_using |>
  mutate(Recipientsex = factor(Recipientsex, levels = c("Male","Female"), labels  = c("Male","Female")),
         Donorsex = factor(Donorsex, levels = c("Female", "Male"), labels  = c("Female", "Male")),
         Smoking = factor(Smoking, levels = c("Yes", "No"), labels  = c("Yes", "No")),
         TypecadavericDBDDCD = factor(TypecadavericDBDDCD, levels = c("Donation after circulatory death", "Donation after brain death"), labels  = c("DCD", "DBD"))
         )
```

###  summarize of data 2: to get the variable description
```{r, echo=TRUE}
# Create a correlation matrix
cor_matrix <- cor(df_using[, sapply(df_using, is.numeric)])

# Plot the correlation matrix
corrplot(cor_matrix, method = "color", type = "upper", 
         tl.col = "black", tl.srt = 45, 
         addCoef.col = "black", number.cex = 0.7)

results2 <- data.frame(
  Variable = character(),
  mean_val = numeric(),
  medium_val = numeric(),
  sd_val = numeric()
)


# Function to create bar plots for factor columns
create_barplot <- function(df, var) {
   ggplot(df, aes_string(x = var, fill = var)) +
    geom_bar() +
    labs(title = paste("Bar Plot of", var), x = var, y = "Proportion") +
    theme_minimal() +
    theme(legend.position = "bottom")
}

# Loop through each column in the data frame
for (var in colnames(df_using)) {
  if (is.numeric(df_using[[var]])){
    results2 <- rbind(results2, data.frame(
      Variable = var,
      mean_val = mean(df_using[[var]]),
      median_val = median(df_using[[var]]),
      sd_val =  sd(df_using[[var]])
    ))
  } else if (!is.numeric(df_using[[var]]) & !inherits(df_using[[var]], "Date")) {
    print(create_barplot(df_using, var))
  }
}

# Print the results data frame
print(results2)
```





## recensoring data because of not accounting for completing risks: only focusing on graft loss
```{r, echo=TRUE}
df_using$status3 = ifelse(df_using$status == "graftloss", 1, 0)
sf = survfit(Surv(time,status3) ~ 1, data=df_using)
plot(sf$time, 1 - sf$surv, xlab = "Days in kidney transplation",
     ylab = "Cumulative distribution function", main = "The chance to experience graftloss before time t")
```
## Exploring the liner/non-linear relationship between the event of "graftloss" and other variables one-by-one

```{r, echo=TRUE}
create_EDA <- function(df, var){
  if (is.numeric(df[[var]])){
    print(var)
    model <- coxph(Surv(time, status3) ~ pspline(df[[var]], df = 4), data = df)
    termplot(model, term = 1, se = TRUE, transform.x = TRUE, partial.resid = TRUE,
             col.term = "blue", col.se = "red", xlabs = paste(var), 
             ylabs = paste("Parietal for Pspline (", var, ")"),
             main = paste("Effect of transformed", var, "on Hazard Function"))
  } else if (is.factor(df[[var]])){
    print(var)
    model1 <- coxph(Surv(time, status3) ~ df[[var]], data = df)
    print(summary(model1))
  }
}

par(mfrow = c(3, 4))
# Loop through each column in the data frame
for (var in colnames(df_using)[c(-25, -26, -27, -28)]){
  create_EDA(df_using, var)
}
```

```{r, echo=TRUE}
library(rstanarm)

```
